---
title: "Survey Analysis: Student Attitudes Towards AI-Generated Content"
subtitle: "Exploring University Students' Trust, Attitudes and Behavioural Responses to AI-Generated Content"
author: "Michelle Nyoni (3519339), MSc Artificial Intelligence, University of Stirling"
date: "`r format(Sys.Date(), '%d %B %Y')`"
output:
  pdf_document:
    toc: true
    number_sections: true
    fig_caption: true
    latex_engine: pdflatex
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    theme: readable
geometry: margin=2.5cm
fontsize: 11pt
urlcolor: blue
---

```{r setup, include = FALSE}
# This chunk sets the defaults for the whole document. include = FALSE means
# the chunk runs but neither the code nor its output appears in the report.
#
#   echo    = TRUE   show the code as well as the results
#   warning = FALSE  hide warnings (see the note in section 6 about ties)
#   message = FALSE  hide startup messages
#   comment = ""     drop the "##" R normally puts before output
#
# options(width = 78) keeps wide output inside the PDF page margins.

knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      comment = "", fig.align = "center",
                      out.width = "100%")
options(width = 78)
```

# How to use this file

Open this file in RStudio. The data file `survey_clean_R.csv` must sit in the
**same folder**. Before anything else choose
**Session -> Set Working Directory -> To Source File Location**.

Then either:

* press **Knit** to produce the whole report as a PDF, or
* run one chunk at a time using the small green arrow at its top right.

## What you need installed

| Needed for | Package | Install once with |
|------------|---------|-------------------|
| The analysis itself | none | nothing to install |
| Knitting | `rmarkdown` | `install.packages("rmarkdown")` |
| Knitting to **PDF** | LaTeX | see below |

Every statistical function used below is built into R. Nothing about the
results depends on an add-on package.

PDF output needs a LaTeX installation. The easiest is TinyTeX, which is small
and installs itself. Run these two lines **once** at the R console, then
restart RStudio:

```
install.packages("tinytex")
tinytex::install_tinytex()
```

The second line takes a few minutes. After that, **Knit** produces a PDF.

If the PDF ever refuses to build and you need output quickly, use the small
arrow beside the Knit button and choose **Knit to HTML** instead. Both outputs
are configured at the top of this file and contain identical results.

# Read the data

`read.csv()` opens a comma-separated file and stores it as a *data frame*,
which is R's word for a table of rows and columns.
`stringsAsFactors = FALSE` tells R to leave text as plain text for now, so
that we can decide ourselves which columns should become categories.

```{r read-data}
survey <- read.csv("survey_clean_R.csv", stringsAsFactors = FALSE)
```

If you get "cannot open file", the working directory is wrong. Run `getwd()`
to see where R is looking.

# Check the data before touching it

Never clean data you have not looked at. Expect 48 rows and 48 columns.

```{r check-size}
dim(survey)
nrow(survey)
```

```{r check-names}
names(survey)
```

The table is too wide to print in full, so the first few rows of the first
eight columns are enough to confirm it loaded correctly. The number before the
comma inside the square brackets selects rows, the part after it selects
columns.

```{r check-head}
head(survey[, 1:8])
```

# Clean the data

## Remove accidental spaces around text answers

Microsoft Forms exported some answers with a space on the end, so
`"Arts and Humanities "` and `"Arts and Humanities"` look identical to a human
but count as two different categories in R. `trimws()`, short for "trim white
space", strips spaces from the start and end of every answer.

This corrects an export fault. It does not change anyone's answer.

```{r trim-whitespace}
survey$age_range          <- trimws(survey$age_range)
survey$study_level        <- trimws(survey$study_level)
survey$subject_area       <- trimws(survey$subject_area)
survey$student_status     <- trimws(survey$student_status)

survey$attitude_useful_ai <- trimws(survey$attitude_useful_ai)
survey$difficulty_detecting_ai <- trimws(survey$difficulty_detecting_ai)

survey$confidence_responsible_use <-
  trimws(survey$confidence_responsible_use)
survey$confidence_judging_reliability <-
  trimws(survey$confidence_judging_reliability)
```

## Harmonise the subject area categories

Respondents typed their own subject area, so two pairs of answers mean the
same thing but are written differently.

**These are judgement calls, not facts.** Delete either line to undo it.

* `"law"` is the same subject as `"Law"`, just typed in lower case.
* `"Environmental Sciences"` is treated as part of Stirling's
  `"Biological and Environmental Sciences"` division.

```{r harmonise-subjects}
survey$subject_area[survey$subject_area == "law"] <- "Law"

survey$subject_area[survey$subject_area == "Environmental Sciences"] <-
  "Biological and Environmental Sciences"
```

## Study level is left exactly as recorded

Four respondents described themselves as "Graduate", "Recent graduate",
"Self employed" and "Currently working post graduation" rather than choosing a
study level. These are reported exactly as given.

**Consequence:** study level has seven categories, four of which contain a
single person. A group of one carries no statistical information, so study
level is reported descriptively only and is never used as a grouping variable
in a test. Section 8 repeats the main tests without these four people.

## Turn the rating questions into ordered factors

An *ordered factor* tells R that these answers have a running order:
"Strongly disagree" sits below "Disagree", which sits below "Neutral", and so
on. Without this, R would sort them alphabetically, which is meaningless here.

The two scales use different wording, so they need separate level lists.

```{r ordered-factors}
agreement_levels <- c("Strongly disagree", "Disagree", "Neutral",
                      "Agree", "Strongly agree")

confidence_levels <- c("Extremely not confident",
                       "Somewhat not confident",
                       "Neutral",
                       "Somewhat confident",
                       "Extremely confident")

survey$attitude_useful_ai <- factor(
  survey$attitude_useful_ai,
  levels = agreement_levels, ordered = TRUE)

survey$confidence_responsible_use <- factor(
  survey$confidence_responsible_use,
  levels = confidence_levels, ordered = TRUE)

survey$difficulty_detecting_ai <- factor(
  survey$difficulty_detecting_ai,
  levels = agreement_levels, ordered = TRUE)

survey$confidence_judging_reliability <- factor(
  survey$confidence_judging_reliability,
  levels = agreement_levels, ordered = TRUE)
```

Every one of these must print `0`. Anything above 0 means an answer was
spelled differently from the level lists above and has been turned into `NA`.

```{r check-factors}
sum(is.na(survey$attitude_useful_ai))
sum(is.na(survey$confidence_responsible_use))
sum(is.na(survey$difficulty_detecting_ai))
sum(is.na(survey$confidence_judging_reliability))
```

## Make numeric copies of the rating questions

The tests further down need numbers, not words. Because the factors above are
ordered, `as.numeric()` converts each answer to its position on the scale:
1 = Strongly disagree up to 5 = Strongly agree.

**These numbers record rank order only.** We are not claiming the gap between
"Agree" and "Strongly agree" is the same size as the gap between "Neutral" and
"Agree". That is exactly why the tests further down are rank-based.

```{r numeric-copies}
survey$attitude_num    <- as.numeric(survey$attitude_useful_ai)
survey$responsible_num <- as.numeric(survey$confidence_responsible_use)
survey$detect_num      <- as.numeric(survey$difficulty_detecting_ai)
survey$reliability_num <- as.numeric(survey$confidence_judging_reliability)
```

## Create the discipline grouping variable

Computing students are compared against everyone else. `ifelse()` reads as
"if the test is true give the first answer, otherwise give the second".
Expect 17 Computing and 31 Other.

```{r discipline}
survey$discipline <- ifelse(
  survey$subject_area == "Computing/ Engineering/ Technology",
  "Computing", "Other")

table(survey$discipline)
```

## Check the tick-box columns hold only 0 and 1

`range()` prints the smallest and largest value found. Both should be 0 and 1.

```{r check-binary}
range(survey[, c("encountered_text", "encountered_images",
                 "encountered_audio", "encountered_videos",
                 "encountered_deepfakes", "encountered_social_media",
                 "encountered_academic", "encountered_unsure")])

range(survey[, c("use_writing", "use_coding", "use_study_support",
                 "use_summarising", "use_brainstorming", "use_images",
                 "use_presentations", "use_entertainment",
                 "use_none", "use_other")])
```

# Descriptive statistics

`table()` counts how many people gave each answer. `prop.table()` turns those
counts into proportions; multiplying by 100 and rounding to one decimal place
gives readable percentages. Read these lines from the inside out.

## Who took part

```{r demographics-age}
table(survey$age_range)
round(100 * prop.table(table(survey$age_range)), 1)
```

```{r demographics-level}
table(survey$study_level)
```

```{r demographics-subject}
table(survey$subject_area)
round(100 * prop.table(table(survey$subject_area)), 1)
```

```{r demographics-status}
table(survey$student_status)
round(100 * prop.table(table(survey$student_status)), 1)

table(survey$interview_willing)
```

## The four rating questions

For each: the counts, the percentages, and the median. The **median** is the
middle answer once everyone is lined up in order. It is the right summary for
a rating scale; the mean is not, because it assumes the gaps between scale
points are equal, which we cannot know.

```{r rating-attitude}
# "AI-generated content can be useful for students"
table(survey$attitude_useful_ai)
round(100 * prop.table(table(survey$attitude_useful_ai)), 1)
median(survey$attitude_num)
```

```{r rating-responsible}
# "How confident do you feel using AI-generated content responsibly?"
table(survey$confidence_responsible_use)
round(100 * prop.table(table(survey$confidence_responsible_use)), 1)
median(survey$responsible_num)
```

```{r rating-detect}
# "I find it difficult to tell whether some content has been generated by AI"
table(survey$difficulty_detecting_ai)
round(100 * prop.table(table(survey$difficulty_detecting_ai)), 1)
median(survey$detect_num)
```

```{r rating-reliability}
# "I feel confident judging whether AI-generated content is reliable"
table(survey$confidence_judging_reliability)
round(100 * prop.table(table(survey$confidence_judging_reliability)), 1)
median(survey$reliability_num)
```

## The four multi-select questions

Each answer option is its own 0/1 column, so one person can appear in several
of them. `colSums()` adds up each column, which counts how many people ticked
it. Percentages are out of all 48 respondents, so they add to more than 100.
That is correct, not an error.

```{r multiselect-encountered}
encountered_columns <- c(
  "encountered_text", "encountered_images", "encountered_audio",
  "encountered_videos", "encountered_deepfakes",
  "encountered_social_media", "encountered_academic",
  "encountered_unsure")

encountered_counts <- sort(colSums(survey[, encountered_columns]),
                           decreasing = TRUE)
encountered_counts
round(100 * encountered_counts / nrow(survey), 1)
```

```{r multiselect-use}
use_columns <- c(
  "use_writing", "use_coding", "use_study_support", "use_summarising",
  "use_brainstorming", "use_images", "use_presentations",
  "use_entertainment", "use_none", "use_other")

use_counts <- sort(colSums(survey[, use_columns]), decreasing = TRUE)
use_counts
round(100 * use_counts / nrow(survey), 1)
```

```{r multiselect-useful}
useful_columns <- c(
  "useful_brainstorming", "useful_summarising", "useful_concepts",
  "useful_coding", "useful_drafting", "useful_revision",
  "useful_not_academic", "useful_unsure")

useful_counts <- sort(colSums(survey[, useful_columns]),
                      decreasing = TRUE)
useful_counts
round(100 * useful_counts / nrow(survey), 1)
```

```{r multiselect-concerns}
concern_columns <- c(
  "concern_misconduct", "concern_overreliance",
  "concern_critical_thinking", "concern_plagiarism",
  "concern_inaccurate", "concern_guidance", "concern_privacy",
  "concern_none")

concern_counts <- sort(colSums(survey[, concern_columns]),
                       decreasing = TRUE)
concern_counts
round(100 * concern_counts / nrow(survey), 1)
```

## The four count variables

These record how many content types each person had met, how many purposes
they used AI for, and so on. They are genuine counts, so the mean and standard
deviation are appropriate here. `summary()` gives the minimum, lower quartile,
median, mean, upper quartile and maximum.

```{r count-variables}
summary(survey$n_encountered); sd(survey$n_encountered)
summary(survey$n_uses);        sd(survey$n_uses)
summary(survey$n_useful);      sd(survey$n_useful)
summary(survey$n_concerns);    sd(survey$n_concerns)
```

# Figures

Three figures, numbered as they appear in the write-up. Each combines several
panels to save space.

`par(mfrow = c(rows, columns))` arranges the next few plots in a grid.
`par(mar = ...)` sets the margins in the order bottom, left, top, right, which
matters where the labels are long. `las = 2` turns the labels sideways.

Short labels keep the panels readable.

```{r short-labels}
short_agreement <- c("Str. disagree", "Disagree", "Neutral",
                     "Agree", "Str. agree")

short_confidence <- c("Not at all", "Somewhat not", "Neutral",
                      "Somewhat", "Extremely")
```

## Figure 1: the four rating questions

```{r figure-1, fig.width = 9, fig.height = 7.5, fig.cap = "The four rating items (n = 48). Panel (b) uses a confidence scale."}
par(mfrow = c(2, 2), mar = c(7, 4, 3, 1))

barplot(table(survey$attitude_useful_ai),
        names.arg = short_agreement,
        main = "(a) AI content can be useful for students",
        ylab = "Respondents", col = "grey70",
        las = 2, ylim = c(0, 25))

barplot(table(survey$confidence_responsible_use),
        names.arg = short_confidence,
        main = "(b) Confidence using AI responsibly",
        ylab = "Respondents", col = "grey70",
        las = 2, ylim = c(0, 25))

barplot(table(survey$difficulty_detecting_ai),
        names.arg = short_agreement,
        main = "(c) Difficult to tell whether content is AI",
        ylab = "Respondents", col = "grey70",
        las = 2, ylim = c(0, 25))

barplot(table(survey$confidence_judging_reliability),
        names.arg = short_agreement,
        main = "(d) Confident judging whether AI content is reliable",
        ylab = "Respondents", col = "grey70",
        las = 2, ylim = c(0, 25))

par(mfrow = c(1, 1))
```

## Figure 2: purposes and concerns

The raw column names are too ugly for a chart, so these lookup tables pair
each column name with a readable label.

Writing them as a **lookup** rather than a plain list matters.
`concern_counts` is sorted by size and two concerns are tied on 18, so their
order is not guaranteed. Indexing the lookup by `names(concern_counts)` makes
each label follow its own bar wherever that bar ends up.

```{r figure-2, fig.width = 10, fig.height = 5, fig.cap = "Purposes AI has been used for (a) and concerns in academic settings (b). Multi-select."}
use_labels <- c(use_summarising   = "Summarising",
                use_brainstorming = "Brainstorming",
                use_writing       = "Writing",
                use_study_support = "Study support",
                use_images        = "Images",
                use_coding        = "Coding",
                use_entertainment = "Entertainment",
                use_presentations = "Presentations",
                use_none          = "Not used",
                use_other         = "Other")

concern_labels <- c(concern_critical_thinking = "Critical thinking",
                    concern_overreliance      = "Over-reliance",
                    concern_plagiarism        = "Plagiarism",
                    concern_inaccurate        = "Inaccuracy",
                    concern_misconduct        = "Misconduct",
                    concern_guidance          = "Guidance",
                    concern_privacy           = "Privacy",
                    concern_none              = "None")

par(mfrow = c(1, 2), mar = c(9, 4, 3, 1))

barplot(use_counts,
        names.arg = use_labels[names(use_counts)],
        main = "(a) Purposes AI has been used for",
        ylab = "Respondents", col = "grey70",
        las = 2, ylim = c(0, 46))

barplot(concern_counts,
        names.arg = concern_labels[names(concern_counts)],
        main = "(b) Concerns in academic settings",
        ylab = "Respondents", col = "grey70",
        las = 2, ylim = c(0, 46))

par(mfrow = c(1, 1))
```

## Figure 3: breadth of use by attitude and by student status

A boxplot shows the spread of a number across groups. The thick line is the
median, the box holds the middle half of the group, and the whiskers reach
most of the rest. Circles beyond the whiskers are unusually high or low
individuals.

The formula `n_uses ~ attitude_num` reads "n_uses as a function of
attitude_num", the same notation used in the course manual.

```{r figure-3, fig.width = 10, fig.height = 5, fig.cap = "Breadth of AI use by attitude towards usefulness (a) and by student status (b), n = 48."}
par(mfrow = c(1, 2), mar = c(7, 4, 3, 1))

boxplot(n_uses ~ attitude_num, data = survey,
        main = "(a) By attitude towards usefulness",
        xlab = "1 = Strongly disagree ... 5 = Strongly agree",
        ylab = "Purposes used (0-8)", col = "grey80")

boxplot(n_uses ~ student_status, data = survey,
        main = "(b) By student status",
        xlab = "", ylab = "Purposes used (0-8)",
        col = "grey80", las = 2)

par(mfrow = c(1, 1))
```

When the document is knitted, these figures are embedded in the PDF as vector
graphics, so they stay sharp at any size and can be copied straight into the
dissertation.

# Statistical tests

**Why these tests and not others.** The rating questions are *ordinal*. We
know "Agree" ranks above "Neutral", but not that it is exactly one unit above.
Tests such as the t-test and ANOVA do arithmetic on the raw scores, which
assumes the scale points are evenly spaced. We cannot support that assumption,
so **rank-based** tests are used instead. These line all 48 respondents up
from lowest to highest, work with each person's *position* in that queue, and
assume nothing about spacing.

| Test | Question it answers | Relation to the manual |
|------|---------------------|------------------------|
| Spearman correlation | Do two ratings rise and fall together? | The rank version of the correlation in Chapter 9 |
| Kruskal-Wallis | Do three or more groups differ? | The rank version of the ANOVA in Chapter 10 |
| Fisher exact test | Are two yes/no variables linked? | Chapter 11. Chosen over chi-square, which needs about five or more expected people per cell; we have fewer |

**Reading the output.** `rho` is correlation strength from -1 to +1; near 0
means no link, and roughly 0.1 is weak, 0.3 moderate, 0.5 strong. The
`p-value` is how easily the result could have arisen by chance alone; below
0.05 is conventionally treated as worth reporting.

**An expected warning.** R produces *"Cannot compute exact p-value with ties"*
for the Spearman tests. Many people gave the same rating, so their ranks are
tied and R falls back on a very good approximation. The warning is harmless
and is hidden in the knitted report by the `warning = FALSE` setting at the
top of this file.

**Sample size caution.** 48 respondents, recruited by convenience rather than
at random. Everything below is **exploratory**. It describes this sample and
points to hypotheses worth testing properly; it does not establish anything
about students in general.

```{r test-1}
# Does a positive attitude go with wider use?
test1 <- cor.test(survey$attitude_num, survey$n_uses,
                  method = "spearman")
test1
```

```{r test-2}
# Do students who find AI hard to detect feel less able to judge reliability?
test2 <- cor.test(survey$detect_num, survey$reliability_num,
                  method = "spearman")
test2
```

```{r test-3}
# Do the two confidence measures move together?
test3 <- cor.test(survey$responsible_num, survey$reliability_num,
                  method = "spearman")
test3
```

```{r test-4}
# Are more concerned students less positive about AI?
test4 <- cor.test(survey$n_concerns, survey$attitude_num,
                  method = "spearman")
test4
```

```{r test-5}
# Do students who use AI more widely have more concerns?
test5 <- cor.test(survey$n_uses, survey$n_concerns,
                  method = "spearman")
test5
```

```{r test-6}
# Does a positive attitude go with feeling able to use AI responsibly?
test6 <- cor.test(survey$attitude_num, survey$responsible_num,
                  method = "spearman")
test6
```

```{r test-7}
# Does breadth of use differ by student status?
test7 <- kruskal.test(n_uses ~ student_status, data = survey)
test7
```

The test does not say *which* group is higher, so look at the groups too.

```{r test-7-groups}
tapply(survey$n_uses, survey$student_status, median)
tapply(survey$n_uses, survey$student_status, mean)
table(survey$student_status)
```

```{r test-8}
# Does breadth of use differ by discipline?
test8 <- kruskal.test(n_uses ~ discipline, data = survey)
test8
tapply(survey$n_uses, survey$discipline, median)
```

```{r test-9}
# Does the number of concerns differ by discipline?
test9 <- kruskal.test(n_concerns ~ discipline, data = survey)
test9
tapply(survey$n_concerns, survey$discipline, median)
```

```{r test-10}
# Do Computing students use AI for coding more than others?
# table() with two arguments builds a cross-tabulation: rows are discipline,
# columns are whether the person ticked coding (0 = no, 1 = yes).
coding_table <- table(survey$discipline, survey$use_coding)
coding_table

test10 <- fisher.test(coding_table)
test10
```

# Correcting for multiple tests

Running ten tests at the 0.05 threshold means that even if nothing at all were
going on, there would be roughly a 40 per cent chance of at least one
"significant" result purely by luck. The Holm correction raises the bar to
account for this.

Results that stay below 0.05 afterwards are the ones to lead with. Results
that stop being significant should be reported honestly as suggestive only.

`$p.value` pulls the p-value out of a stored test result.

```{r holm-correction}
p_values <- c(test1$p.value, test2$p.value, test3$p.value,
              test4$p.value, test5$p.value, test6$p.value,
              test7$p.value, test8$p.value, test9$p.value,
              test10$p.value)

names(p_values) <- c("1 attitude-uses",
                     "2 detect-reliability",
                     "3 responsible-reliability",
                     "4 concerns-attitude",
                     "5 uses-concerns",
                     "6 attitude-responsible",
                     "7 uses-status",
                     "8 uses-discipline",
                     "9 concerns-discipline",
                     "10 discipline-coding")

round(p_values, 4)                              # before correction
round(p.adjust(p_values, method = "holm"), 4)   # after correction
```

# Sensitivity check

Four respondents are not currently students. This section repeats the main
tests without them, to show whether their presence changed anything.

`%in%` asks "is this value one of the following?" and `!` reverses it, so the
line below reads "keep the rows whose study level is **not** one of these
four". The comma before the closing bracket means "all columns". Expect 44.

```{r sensitivity-subset}
non_student_levels <- c("Currently working post graduation",
                        "Self employed", "Graduate", "Recent graduate")

students_only <- survey[!(survey$study_level %in% non_student_levels), ]

nrow(students_only)
```

```{r sensitivity-tests}
cor.test(students_only$attitude_num, students_only$n_uses,
         method = "spearman")

cor.test(students_only$detect_num, students_only$reliability_num,
         method = "spearman")

cor.test(students_only$responsible_num, students_only$reliability_num,
         method = "spearman")

cor.test(students_only$n_concerns, students_only$attitude_num,
         method = "spearman")

kruskal.test(n_uses ~ student_status, data = students_only)

fisher.test(table(students_only$discipline, students_only$use_coding))
```

Compare each of these against the matching test in section 6. If the direction
and rough size of each result are unchanged, the findings do not depend on
those four respondents, and that can be stated in the dissertation.

# Session information

Recording the R version makes the analysis reproducible by someone else later.

```{r session-info}
sessionInfo()
```
